---
output: github_document
---

[![CRAN status](https://www.r-pkg.org/badges/version/Rparadox)](https://CRAN.R-project.org/package=Rparadox)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/grand-total/Rparadox)](https://CRAN.R-project.org/package=Rparadox)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/Rparadox)](https://CRAN.R-project.org/package=Rparadox)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/last-week/Rparadox)](https://CRAN.R-project.org/package=Rparadox)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/last-day/Rparadox)](https://CRAN.R-project.org/package=Rparadox)
[![Ask DeepWiki](https://deepwiki.com/badge.svg)](https://deepwiki.com/celebithil/Rparadox)

# Rparadox: A Modern Interface for Reading Paradox Databases in R

**Rparadox** provides a simple and efficient way to read data from Paradox database files (`.db`) directly into R as modern `tibble` data frames. It uses the underlying `pxlib` C library to handle the low-level file format details and provides a clean, user-friendly R interface.

This package is designed to "just work" for the most common use case: extracting the full dataset from a Paradox table, including its associated BLOB/memo file (`.mb`).

------------------------------------------------------------------------

## Features

-   **Direct Reading:** Reads Paradox `.db` files without needing database drivers or external software.
-   **Tibble Output:** Returns data in the `tibble` format, which is fully compatible with the Tidyverse ecosystem.
-   **Automatic BLOB Handling:** Automatically detects, attaches, and reads data from associated memo/BLOB (`.mb`) files.
-   **Encryption Support:** Supports reading password-protected Paradox files.
-   **Character Encoding Control:** Automatically handles character encoding conversion to UTF-8 and allows the user to manually override the source encoding for files with incorrect headers.
-   **Type Conversion:** Correctly maps Paradox data types to their corresponding R types, including `Date`, `Time` (`hms`), `Timestamp` (`POSIXct`), `Logical`, `Integer`, `Numeric`, and binary `blob` objects.

------------------------------------------------------------------------

## Installation

Install the stable version from CRAN:

```{r, eval=FALSE}
install.packages("Rparadox")
```

You can install the development version from GitHub using the `devtools` package.

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("celebithil/Rparadox")
```

------------------------------------------------------------------------

## Usage

See the package vignette for complete usage examples:

```{r, eval=FALSE}
vignette("Rparadox")
```

------------------------------------------------------------------------

## Links

-   pxlib C library: <https://github.com/steinm/pxlib>
-   CRAN page: <https://cran.r-project.org/package=Rparadox>
-   Bug reports: <https://github.com/celebithil/Rparadox/issues>
